The Precambrian paleogeography of Laurentia

### 2020-10-01 Manuscript/Laurentia_Paleogeo_Manuscript.tex
write Siberia connection section and make progress on Australia

There is significantly less preserved Mesoproterozoic crustal growth on the western margin of Laurentia (Fig. \ref{fig:Laurentia_map}), and the  tectonic history through the Mesoproterozoic Era is not as well elucidated as on the southern to eastern margin. The 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with a ca. 1.47 to 1.40 rift \citep{Evans2000c}. While the rift is typically interpreted as being intracontinental \citep{Lydon2004a}, the tectonic setting in which it formed is debated. \citet{Hoffman1989c} proposed that it may be a remnant back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep{Jones2015a}. This region is interpreted to have been subsequently deformed during a ca. 1.36 to 1.33 Ga event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Grenville_Reconstruction.pdf}
\caption{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic}. This reconstruction is modified from \cite{Swanson-Hysell2019a}. The record of paleomagnetic poles implies rapid motion which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}
\label{fig:Grenville_reconstructions}
\end{figure*}

@@

It has long been argued that there are shared aspects of the geologic history between Australia, East Antarctica and Laurentia. Such correlations has led to a number of different proposed connections that place Australia + East Antarctica...

@@

%\subsubsection{Siberia}

>>>>

There is significantly less preserved Mesoproterozoic crustal growth on the western margin of Laurentia (Fig. \ref{fig:Laurentia_map}) and the tectonic history through the Mesoproterozoic Era is not as well elucidated as on the southern to eastern margin. There are thick successions of Paleoproterozoic sediments such as the ca. 1.66 to 1.62 Ga Wernecke Supergroup in Yukon, Canada that comprises $\sim$14 km of  fine-grained siliciclastic and carbonate rocks\citep{Furlanetto2016a}. This succession is interpreted to have resulted from rifting followed by passive margin thermal subsidence \citep{Furlanetto2016a}. These metasedimentary rocks were metamorphosed during the ca. 1.60 Ga Racklan orogeny associated with collision of an arc terrane and potentially a conjugate continent \citep{Furlanetto2013a, Furlanetto2016a}. Further south along Laurentia's western margin, the 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with a ca. 1.47 to 1.40 rift \citep{Evans2000c}. While the rift is typically interpreted as being intracontinental \citep{Lydon2004a}, the tectonic setting in which it formed is debated. \citet{Hoffman1989c} proposed that it may be a remnant back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep{Jones2015a}. This region is interpreted to have been subsequently deformed during a ca. 1.36 to 1.33 Ga event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Grenville_Reconstruction.pdf}
\caption{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic}. The hypothesized connection between Siberia and Laurentia is implemented here following \citep{Evans2016b}. This reconstruction is modified from \cite{Swanson-Hysell2019a}. The record of paleomagnetic poles implies rapid motion which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}
\label{fig:Grenville_reconstructions}
\end{figure*}

@@

It has long been argued that there are shared aspects of the geologic history between Australia, East Antarctica and Laurentia. Such correlations has led to a number of different proposed connections that place Australia + East Antarctica along the western margin of Laurentia.

Metamorphism ca. 1.6 Ga on the eastern margin of the North Australia craton associated with the Issan-Jana orogeny has been interpreted to the be result of collisional orogenesis with the western Laurentia margin \citep{Nordsvan2018a, Pourteau2018a}. That Laurentia was the conjugate continent for this orogeny is argued to be supported by detrital zircon date spectra from the Georgetown Inlier of the eastern North Australia craton that have similarities with possible Laurentia sources \citep{Nordsvan2018a}. The model of \cite{Nordsvan2018a} the inlier to have been a continental ribbon that rifted from Laurentia ca. 1.68 Ga and was then caught up in ca. 1.60 Ga collision between the North Australia craton and northwest Laurentia. The ca. 1.60 Ga Racklan-Forward orogeny recorded in northwest Laurentia records arc-continent collision that could have been followed by continent-continent collision \citep{Furlanetto2013a}.

Metamorphism a. 1.60 Ga Paleoproterozoic

@@

\subsubsection{Siberia}

Southern Siberia and northern Laurentia have been proposed to be connected on the basis of a similar history of Paleoproterozoic collision \citep{Rainbird1998a}, matches in the age of mafic intrusives rocks interpreted as shared large igneous provinces from the Paleoproterozoic to the Neoporterozoic \cite{Ernst2016a}, and comparisons of paleomagnetic poles \citep{Evans2011a, Evans2016b}. U-Pb dates from mafic intrusive rocks are interpreted by \cite{Ernst2016a} as resulting from shared large igneous province  between southern Siberia and northern Laurentia from the time of Laurentia's amalgamation ca. 1.8 Ga all the way up to the time of the ca. 720 Ma Franklin large igneous province (LIP). Mafic intrusions and lavas that are grouped as the Irkutsk LIP in Siberia have been dated in Siberia to be similar in age to dates developed from the Franklin LIP \citep{Denyszyn2009a, Ernst2016a}. Comparisons of paleomagnetic poles between Laurentia and Siberia support such a tight and internally stable fit between southern Siberia and northern Laurentia from ca. 1.64 to at least ca. 0.76 Ga  \citep{Evans2016b}. Overlap in Laurentia and Siberia pole positions with such a reconstruction are achieved in the Statherian Period of the Paleoproterozoic (the ca. 1.64 Ga Nersa complex of Siberia compared to the Melville Bugt dikes of Laurentia), the Calymmian Period of the Mesoproterozoic (the ca. 1.50 Ga Anabar intrusions with the ca. 1.48 Ga St. Francois Mountains igneous province of Laurentia), the Stentian Period of the Mesoproterozoic (a number of roughly constrained Siberia poles with the 1.11 to 1.08 Ga Keweenawan Track of Laurentia) and the Tonian Period of the Neoproterozoic (the ca.  Kitoi pole of Siberia with Tonian Laurentia poles).  The correlation of the Malgina pole of Siberia with the Keweenawan Track gains additional support in that it correlates the normal-polarity Maya superchron \citep{Gallet2012a} with the Portage Lake normal-polarity zone \citep{Swanson-Hysell2019a} that is interpreted as a normal-polarity superchron (termed the Keweenawan Normal Superchron in \citealp{Driscoll2016a}). This correlation both works with the tight fit and resolves hemispheric ambiguity to put both cratons together in the northern hemisphere ca. 1100 Ma.

Connections may have existed prior to Laurentia assembly such as the hypothesized connection between the Slave Province and the Tungus Province of Siberia which has the Thelon orogen correlated to Paleoproterozoic orogenesis in the Akitkan fold belt \citep{Condie1994a, Rainbird1998a, Evans2011a}. The emplacement of the Franklin LIP ca. 720 Ma could have initiated break-up of a joint Laurentia-Siberia continent ending the long Proterozoic connection between the cratons.
